### 2023-12-27 ms.qmd
Updates

Cognition is defined as the processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision making (@shettleworth). These cognitive processes underpin several aspects of animals' ecology such as foraging, mate choice, antipredatory strategies, and/or social behaviours, all crucial for the survival and reproduction of animals (@dukas_evolutionary_2004). Notably, learning - the acquisition of neuronal representations of new information (@dukas_evolutionary_2004) - is seen as fundamental for coping with environmental changes by enabling individuals to create new associations between events (@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013). However, the capacity of individuals to acquire new information exhibits natural variation influenced by factors like age, sex, gut microbiota, or the environment where animals develop (@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014; @alemohammad_2022_microbiota_learning). The developmental environment, in particular, plays a pivotal role, as the brain is especially susceptible to environmental influences during early stages of development (@zhu_prenatal_2004). 
Prenatal Glucocorticoids (GCs) and prenatal thermal environment are known to play a prominent role in shaping learning abilities in different taxa (see @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @crino_corticosterone_2014; @abayarathna_effects_2020). GCs - hormones related to organisms' response to stress (@sapolsky_how_2000) - exert sustained effects on neural structure and physiology that are associated with animals' performance on learning tasks (@lemaire_prenatal_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016). Some studies have demonstrated that prenatal stress and high prenatal GC levels impair with learning (@lemaire_prenatal_2000; @zhu_prenatal_2004; @farrell_developmental_2015-learn), while others showed diverse effects depending on factors such as sex or the nature of the learning task (@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016). 
Similarly, some experiments have shown significant impacts of prenatal temperature on learning in ectotherms (@amiel_hotter_2012; @amiel_egg_2014;). For instance, high incubation temperatures have been linked with faster learning rates in skinks (@amiel_hotter_2012; @amiel_egg_2014), while velvet geckos incubated at temperatures over their natural range learn slower than those incubated within the natural thermal limits (@abayarathna_effects_2020). In this vein, the effect of prenatal temperature appears to be linked to alterations in neural structure and metabolic activity (@coomber_independent_1997; @sakata_neural_2000; @amiel_effects_2017; @beltran_are_2021) that are similar to those resulting from prenatal increased stress or GC levels (@lemaire_prenatal_2000; @zhu_prenatal_2004; @du_dynamic_2009). This suggests that prenatal GCs and temperature can act on the same physiological mechanisms and, thus, both could interact to shape individual variation in learning skills (@amiel_effects_2017; @noble_developmental_2018). Furthermore, GCs can play an pivotal role in determining vertebrate responses to elevated temperatures (@Crino_2023) potentially fostering natural interactions between temperature and GCs. Despite the proximate similarities of prenatal GCs and temperature effects and the potential role of GCs in vertebrates response to elevated temperatures, our understanding of how these two factors interact remains incomplete.

@@

_L. guichenoti_ and _L. delicata_ are small (∼35--55 mm snout-vent length (SVL)), oviparous, and generalist skinks that usually share the same habitat in suburban areas throughout south-eastern Australia ([@chapple_know_2011]). They also have similar breeding periods, but with some differences in reproductive output: while *L. delicata* lays 1 to 6 eggs in only one clutch per season, *L. guichenoti* clutches are smaller (1-5 eggs per clutch) but they usually lay two clutches per season ([@chapple_know_2011]; [@chapple_biology_2014]).

@@

Two weeks before we started the training phase (see below), lizards were moved to the experimental arena  for acclimatation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the other. These new enclosures were placed in two rooms in 7 different racks associated to 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). The number of lizards per species and treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatation and all the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied *ad libitum.* We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The rooms temperature was set to between 22-24 Celsius. We tested temperatures with laser temperature guns to ensure such a set up established a thermal gradient and we there were no significant differences between treatments or species (see Supplementary Material).

### CORT and Temperature manipulation

To test empirically the effect of early environment we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes ('Cold' -- 23ºC ± 3ºC or 'Hot' -- 30ºC ± 3ºC) in a 2x2 factorial design (Fig. 1a). We first allocated eggs to one of two different treatments: in one of the treatments, eggs were topically supplied with 5µL of CORT dissolved in 70% Ethanol and 30% DMSO (vehicle) at a final (10 pg CORT/mL) concentration (CORT treatment); while eggs in the other group (Control treatment) received an equal volume of the vehicle. CORT concentration employed in the CORT treatment represents 2 standard deviations above the mean natural concentration obtained in eggs from both species (non-published data). Then, eggs were incubated in one of the two previously mentioned temperature regimes ('Cold' or 'Hot') until hatching. The number of eggs per clutch assigned to each hormone and temperature treatment were counterbalanced in both species.

@@

To estimate learning skills, we tested skinks' ability to locate a food reward in a series of behavioural tasks (Fig. 1b) (see [@leal_behavioural_2012]; [@clark_colour_2014]). First, we performed a training phase where lizards had to learn to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm) identical to the ones from the experiment except for the colour (see below). We divided this training phase into three stages (see Fig. 1b): in the first stage, lizards had to eat a small, frozen cricket (*A. domestica*) from an opaque petri dish (3D x 1.6H cm) placed in the middle of their enclosure; in the second stage, the petri dish with the cricket was placed on top of the white 3D printed ramps; and finally, the cricket was left inside a well (3D x 1.75H cm) on the top of the ramp in the third and last stage. Every trial began when we left the feeding block (petri dish, ramp, or both) inside the enclosure and finished one hour later when we took it away. At the end of the experiment, we recorded whether the cricket has been consumed or not. Trial was considered successful if the lizard could locate and consume the reward, while completion of each stage required the lizards to eat the crickets in at least 5 out of 6 trials. This phase lasted 38 days until all the lizards were able to eat from the ramp; only in one case we decided not to use the lizard because its behaviour was not consistent over the course of the training phase (see Supplementary material).

In the next phase, we trained lizards to associate between colour and a food reward (colour association task in Fig. 1b). The test was like the third stage of the training phase, but here lizards were presented with three feeders that differed in the colour. We placed the food reward (small, frozen, *A. domestica* crickets) inside the wells of the three feeders, covering two of the crickets with 3D-printed lids (3D x 0.5H cm) so prey was only accessible in one ramp. The food reward was placed in all three wells to avoid the use of chemical cues to determine the correct choice; our lids had a series of small holes on the top for the same reason. The colours of the feeders were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours (REF). To control for potential colour preference that could bias our results, we split the subjects in two groups counterbalanced by treatment and species; in one group the correct choice (i.e., the ramp with non-covered cricket) was the blue one in the associative task and red in the reversal, while we did the opposite for the other half. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in this task once a day for 35 days.

After the colour association phase, we performed a choice reversal task. This task was like the colour association test, except that the attainability of prey was indicated by a different colour, requiring the lizards to form a novel association between the new colour and the food reward. This test was done once a day for 40 days.

All trials were done daily between the 6^th^ of March until the 26^th^ of June 2023, beginning at 11 am until 12 pm. Trials in the learning phases (colour associative task and reversal tasks) were recorded with different CCTV systems always using the same camera per individual. Videos were analysed manually using a standard video player (IINA) by PR, who recorded whether the first choice made by each subject was the correct feeder or not.

```{r, Fig1}
#| label: fig-fig1
#| fig-cap: ""
knitr::include_graphics("./Others/LEARN_FIG_1.svg") 
```

>>>>

Cognition is defined as the processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision making (@shettleworth). These cognitive processes underpin several aspects of animals' ecology such as foraging, mate choice, antipredatory strategies, and/or social behaviours, that are crucial for the survival and reproduction of animals (@dukas_evolutionary_2004). Particularly, learning - the acquisition of neuronal representations of new information (@dukas_evolutionary_2004) - is seen as fundamental for coping with environmental changes by enabling individuals to create new associations between events (@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013). However, the capacity of individuals to acquire new information exhibits natural variation influenced by factors like age, sex, gut microbiota, or the environment where animals develop (@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014; @alemohammad_2022_microbiota_learning). The developmental environment, in particular, plays a pivotal role, as the brain is especially susceptible to environmental influences during early stages of development (@zhu_prenatal_2004). Hence, investigating the effects of the developmental environment on learning can be essential to understand the evolution of learning and predict animals' responses towards environmental change.
In this sense, prenatal Glucocorticoids (GCs) and prenatal thermal environment are known to play a prominent role in shaping learning abilities in different taxa (see @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @crino_corticosterone_2014; @abayarathna_effects_2020). GCs - hormones related to organisms' response to stress (@sapolsky_how_2000) - exert sustained effects on neural structure and physiology that are associated with animals' performance on learning tasks (@lemaire_prenatal_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016). Some studies have demonstrated that prenatal stress and high prenatal GC levels impair with learning (@lemaire_prenatal_2000; @zhu_prenatal_2004; @farrell_developmental_2015-learn), while others showed diverse effects depending on factors such as sex or the nature of the learning task (@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016). Similarly, some experiments have shown significant impacts of prenatal temperature on learning in ectotherms (@amiel_hotter_2012; @amiel_egg_2014; @dayananda_incubation_2017; @abayarathna_effects_2020). For instance, high incubation temperatures have been linked with faster learning rates in skinks (@amiel_hotter_2012; @amiel_egg_2014), while velvet geckos incubated at temperatures over their natural range learn slower than those incubated within the natural thermal limits (@abayarathna_effects_2020). In this vein, the effect of prenatal temperature appears to be linked to alterations in neural structure and metabolic activity (@coomber_independent_1997; @sakata_neural_2000; @amiel_effects_2017; @beltran_are_2021) that share some similarities with those resulting from prenatal increased stress or GC levels (@lemaire_prenatal_2000; @zhu_prenatal_2004; @du_dynamic_2009). This suggests that prenatal GCs and temperature can act on the same physiological mechanisms and, thus, both could interact to shape individual variation in learning skills (@noble_developmental_2018). Furthermore, GCs can play an pivotal role in determining vertebrate responses to elevated temperatures (@Crino_2023) potentially fostering natural interactions between temperature and GCs. Despite the proximate similarities of prenatal GCs and temperature effects and the potential role of GCs in vertebrates response to elevated temperatures, our understanding of how these two factors interact remains incomplete.

@@

_L. guichenoti_ and _L. delicata_ are small (∼35--55 mm snout-vent length (SVL)), oviparous, and generalist skinks that usually share the same habitat in suburban areas throughout south-eastern Australia (@chapple_know_2011). They also have similar breeding periods, but with some differences in reproductive output: while *L. delicata* lays 1 to 6 eggs in only one clutch per season, *L. guichenoti* clutches are smaller (1-5 eggs per clutch) but they usually lay two clutches per season (@chapple_know_2011; @chapple_biology_2014).

@@

Two weeks before we started the training phase (see below), lizards were moved to the experimental arena for acclimatation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the other. These new enclosures were placed in two rooms in 7 different racks associated to 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). The number of lizards per species and treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatation and all the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied *ad libitum*. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The rooms temperature was set to between 22-24 Celsius.

### CORT and Temperature manipulation
To test empirically the effect of early environment we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes ('Cold' -- 23ºC ± 3ºC or 'Hot' -- 30ºC ± 3ºC) in a 2x2 factorial design (Fig. 1a). We first allocated eggs to one of two different treatments: CORT treatment, where eggs were topically supplied with 5µL of CORT dissolved in 70% Ethanol and 30% DMSO (vehicle) at a final (10 pg CORT/mL) concentration (CORT treatment); and a Control treatment, where eggs received an equal volume of the vehicle. CORT concentration employed in the CORT treatment represents 2 standard deviations above the mean natural concentration obtained in eggs from both species (non-published data). Then, eggs were incubated in one of the two previously mentioned temperature regimes ('Cold' or 'Hot') until hatching. The number of eggs per clutch assigned to each hormone and temperature treatment were counterbalanced in both species.

@@

To estimate learning skills, we tested skinks' ability to locate a food reward in a series of behavioural tasks (Fig. 1b) (see @leal_behavioural_2012; @clark_colour_2014). First, we performed a training phase where lizards had to learn to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm) identical to the ones from the experiment except for the colour (see below). We divided this training phase into three stages: in the first stage, lizards had to eat a small, frozen cricket (*A. domestica*) from an opaque petri dish (3D x 1.6H cm) placed in the middle of their enclosure (Fig. 1B, Stage 1); in the second stage, the petri dish with the cricket was placed on top of the white 3D printed ramps (Fig. 1B, Stage 2); and finally, the cricket was left inside a well (3D x 1.75H cm) on the top of the ramp in the third and last stage (Fig. 1B, Stage 3). Every trial began when we left the feeding block (petri dish, ramp, or both) inside the enclosure and finished one hour later when we took it away. At the end of the experiment, we recorded whether the cricket has been consumed or not. Trial was considered successful if the lizard could locate and consume the reward, while completion of each stage required the lizards to eat the crickets in at least 5 out of 6 trials. This phase lasted 38 days until all the lizards were able to eat from the ramp; only in one case we decided not to use the lizard because its behaviour was not consistent over the course of the training phase.
In the next phase, we trained lizards to associate between colour and a food reward (Associative task in Fig. 1b). The test was like the third stage of the training phase, but here lizards were presented with three feeders that differed in the colour. We placed the food reward (small, frozen, *A. domestica* crickets) inside the wells of the three feeders, covering two of the crickets with 3D-printed lids (3D x 0.5H cm) so prey was only accessible in "the correct" ramp. The food reward was placed in all three wells to avoid lizard using prey chemical cues, and the lids had a series of small holes on the top to allow the release of those chemicals. The colours of the feeders were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours (@Baden_Osorio_2019_Vert_vision). To control for potential colour preference that could bias our results, we split the subjects into two groups counterbalanced by treatment and species: in one group the correct choice (i.e., the ramp with non-covered cricket) was the blue one in the associative task and red in the reversal (see below), while we did the opposite for the other group. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in this task once a day for 35 days.
After the colour association phase, we performed a choice reversal task (Reversal task in Fig. 1B). This task was like the colour association test, except that the attainability of prey was indicated by a different colour, requiring the lizards to form a novel association between the new colour and the food reward. This test was done once a day for 40 days.
All trials were done daily between the 6^th^ of March until the 26^th^ of June 2023, beginning at 11 am until 12 pm. Trials in the learning phases (colour associative task and reversal tasks) were recorded with different CCTV systems always using the same camera per individual. Videos were analysed manually using a standard video player (IINA) by PR, who recorded whether the first choice made by each subject was the correct feeder or not. A choice was considered to be made if the head of the lizard was inside the well of one of the ramps. PR was blinded to the treatments of the lizards during the analyses of the video. We considered a trial failed if there was no choice in one hour of recording and those trials were considered as 'non data' in the analyses.

```{r, Fig1}
#| label: fig-fig1
#| fig-cap: "Experimental design of early environment manipulation (**A**) and learning tasks (**B**). Stages 1-3 indicate the different phases of the habituation process. In the associative and reversal tasks, white lids indicate the ramps where the food reward was not attainable."
knitr::include_graphics("./Others/LEARN_FIG_1.svg") 
```
